Determinism, Complexity, and Predictability in Computer Performance

### 2014-03-04 methods.tex
edits to intro,related and methods
editing

\section{Experimental Methods and Time Series Explanations}

 \begin{enumerate}
 \item \cmark  Experimental methods: (how we collect the time series and what the times series areThis should be HPM PAPI, which programs we model
\item \cmark program descriptions and relevant citations
\subitem \cmark description of \col
\subitem \cmark description of \gcc
\subitem \cmark description of \svd 
\subitem \cmark description of svd regimes
 \end{enumerate}

\subsection{Time Series Collection}

@@

Obviously using a system to measure itself can cause interference--i.e, are you measuring the generating process cleanly or are you adding noise to the process---but due diligence was put into monitoring the generating process of these time series without interfering with it in a significant way. For an in-depth explanation of this custom-measurement infrastructure as well as discussion on choice of interrupt rate see 
\cite{zach-IDA10,mytkowicz09,todd-phd}.

\subsection{The Programs}
We collect performance traces of processor efficiency (IPC) of three exemplar programs---a simple microkernel(\col) and two complex programs: one from the
SPEC 2006CPU benchmark suite (\gcc), and one from LAPACK (\svd). In this section we provide a quick explanation of each program as well as an example performance trace (Seen in Figures \ref{fig:col-ts}-\ref{fig:svd-ts-colored}.
\subsubsection{\col}
\col is a simple three-line C program that repeatedly initializes the upper triangle of a  2048 X 2048 matrix in column-major order. While this is a very simple program it has been shown to produce time series with very complicated behavior. In fact, in \cite{mytkowicz09} it was shown that time series generated from this simple program can exhibit everything from periodicity to deterministic chaos.  A time series of the processor efficiency (IPC) of this program is shown in Figure \ref{fig:col-ts}.

@@

\subsubsection{\gcc}

\gcc is a benchmark program included in the SPEC CPU2006 benchmark suite \cite{spec}. This program is written by Richard Stallman and is based on Version 3.2 of {\tt gcc}. This benchmark essentially runs as a compiler with many of its optimization flags enabled, compiling a series of input files and generating x86-64 assembly code files intended for execution on AMD Opteron processors\cite{spec}. This program is more typical of programs that are used every day, as opposed to the simple micro-kernel, \col. A time series of the processor efficiency of this program running on the Intel i7-2600 is shown in Figure \ref{fig:gcc-ts}.

@@

\subsubsection{\svd}

\svd is a Fortran program from the LAPACK suite \cite{lapack} which calculates the singular value decomposition (SVD) of a rectangular $M$ by $N$  real-valued matrix which need not contain special structure such as symmetric, positive definite etc. To execute this program we input a 750 x 1000 matrix of randomly generated entries\footnote{Multiple randomly generated matrices were investigated but no measurable effect was present in the resulting time series.}. A ``regime"-colored time series of the IPC of this program is shown in Figure \ref{fig:svd-ts-colored}. We now describe why this is done.

@@

This is caused by the code of \svd moving between different subroutines as the SVD algorithm evolves, something not present in either \gcc or \col. We call these changes in \svd dynamics \emph{\svd regimes} and we have colored each of the six regimes differently in Figure \ref{fig:svd-ts-colored}. The advantage to splitting \svd into these regimes is to explore how complexity and predictability evolve over time for a system in drift. In Section \ref{sec:wpeRegime} we explore and validate the choice of these visually selected regime windows extending techniques from \cite{cao2004det}. Note, the purpose of this paper is \emph{not} to rigorously explore regime detection but to explore quantifying complexity of a time series. As such, making these regime splits serves to provide an additional 90 unique time series\footnote{Six Regimes with 15 individual runs each.} to explore. For notational convenience we refer to these time series as {\tt dgesdd$_i$} with $i \in \{1\dots6\}$ where $i$ corresponds to a regime of \svd. The regimes are labeled from left to right.

%\begin{figure}[htbp]
%  \centering
%  \begin{subfigure}[t]{\textwidth}
%  \centering
%    \includegraphics[width=0.75\textwidth]{figs/%colFullTS}
%    \caption{\col Time Series}
%    \label{fig:col-ts}
%  \end{subfigure}%
%  \\

%  \begin{subfigure}[t]{\textwidth}
%  \centering
%    \includegraphics[width=0.75\textwidth]{figs/%gccfullts}
%    \caption{\gcc Time Series}
%    \label{fig:gcc-ts}
%  \end{subfigure}
%    \begin{subfigure}[t]{\textwidth}
%    \centering
%    \includegraphics[width=0.75\textwidth]{figs/%SVD1RegimesColored}
%    \caption{\svd Time Series with Regime Coloring}
%    \label{fig:svd-ts-colored}
%  \end{subfigure}
%  \caption{[[Maybe this should be just a picture of %each time series and then have another figure with %color after explaining heuristically the reigmes. In %(a) the instructions executed per CPU clock cycle%
%    (IPC) during the execution of \col. Each point is the average IPC during a 100,000
%    instruction period. Similarly in (b) is a time %series of the IPC during the execution of \gcc.}%\label{fig:sample-ts}%
%    \end{figure}

>>>>

\section{Experimental Methods and Time Series Explanations}\label{sec:methods}

% \begin{enumerate}
% \item \cmark  Experimental methods: (how we collect the %time series and what the times series areThis should be HPM %PAPI, which programs we model
%\item \cmark program descriptions and relevant citations
%\subitem \cmark description of \col
%\subitem \cmark description of \gcc
%\subitem \cmark description of \svd 
%\subitem \cmark description of svd regimes
% \end{enumerate}

\subsection{Time Series Collection {\color{blue} EDITABLE}}

@@

Obviously using a system to measure itself can cause interference--i.e, are you measuring the generating process cleanly or are you adding noise to the process?---but due diligence was put into monitoring the generating process of these time series without interfering with it in a significant way. For an in-depth explanation of this custom-measurement infrastructure as well as discussion on choice of interrupt rate see 
\cite{zach-IDA10,mytkowicz09,todd-phd}.

\subsection{The Programs: [[Maybe The generating processes]]{\color{blue} EDITABLE}}
We collect performance traces of processor efficiency (IPC) of three exemplar programs, [[If we change the title of the section, add this: which act as our generating processes]]---a simple microkernel(\col) and two complex programs: one from the
SPEC 2006CPU benchmark suite (\gcc), and one from LAPACK (\svd). In this section we provide a quick explanation of each program as well as an example performance trace (Seen in Figures \ref{fig:col-ts}-\ref{fig:svd-ts-colored}.
\subsubsection{\col {\color{blue} EDITABLE}}
\col is a simple three-line C program that repeatedly initializes the upper triangle of a  2048 X 2048 matrix in column-major order. While this is a very simple program it has been shown to produce time series with very complicated behavior. In fact, in \cite{mytkowicz09} it was shown that time series generated from this simple program can exhibit everything from periodicity to deterministic chaos.  A time series of the processor efficiency (IPC) of this program is shown in Figure \ref{fig:col-ts}.

@@

\subsubsection{\gcc {\color{blue} EDITABLE}}

\gcc is a benchmark program included in the SPEC CPU2006 benchmark suite \cite{spec}. This program is written by Richard Stallman and is based on Version 3.2 of {\tt gcc}. This benchmark essentially runs as a compiler with many of its optimization flags enabled, compiling a series of input files and generating x86-64 assembly code files intended for execution on AMD Opteron processors\cite{spec}. This program is more typical of programs that are used every day, as opposed to a simple micro-kernel,e.g., \col. A time series of the processor efficiency of this program running on the Intel i7-2600 is shown in Figure \ref{fig:gcc-ts}.

@@

\subsubsection{\svd {\color{blue} EDITABLE}}

\svd is a Fortran program from the LAPACK suite \cite{lapack} which calculates the singular value decomposition (SVD) of a rectangular $M$ by $N$  real-valued matrix which need not contain special structure such as symmetric, positive definite etc. To execute this program we input a 750 x 1000 matrix of randomly generated entries\footnote{Multiple randomly generated matrices were investigated but no measurable effect was present in the resulting time series.}. A ``regime-colored" time series of the IPC of this program is shown in Figure \ref{fig:svd-ts-colored}. We now describe why this is done.

@@

This is caused by the code of \svd moving between different subroutines as the SVD algorithm evolves, something not present in either \gcc or \col. We call these changes in \svd dynamics \emph{\svd regimes} and we have colored each of the six regimes differently in Figure \ref{fig:svd-ts-colored}. The advantage to splitting \svd into these regimes is to explore how complexity and predictability evolve over time for a system in drift. In Section \ref{sec:wpeRegime} we explore and validate the choice of these visually selected regime windows extending techniques from \cite{cao2004det}. Note, the purpose of this paper is \emph{not} to rigorously explore regime detection but to explore quantifying complexity of a time series. As such, making these regime splits serves to provide an additional 90 unique time series\footnote{Six Regimes with 15 individual runs each.} to explore. For notational convenience we refer to these time series as {\tt dgesdd$_i$} with $i \in \{1\dots6\}$ where $i$ corresponds to a regime of \svd, ordered from left to right.

### 2014-03-04 model.tex
editing

\section{Modeling}\label{sec:model}

@@

\subsection{Primitive Strategies to combat complexity: Random Walk and \naive}\label{sec:simple}
In this section we cover two very simple forecast models: Random Walk and \naive. With a Random Walk predictor to forecast at time $i$ we simply use the last observed measurement: $$p_i = X_{i-1,obs}$$. The \naive  prediction strategy is to forecast the rolling average up to that point. i.e., the prediction $p_i$ is simply the average of all observed points up to that point: $$p_i = \sum_{j=1}^{i-1}\frac{X_{j,obs}}{i-1}$$.

While both of these methods are simplistic they are not without merit. When time series possess very little predictive structure,i.e., they exist on the very top of the complexity spectrum these two methods can sometimes be \emph{on-average} very good choices for a forecasting method, as we will see in our results. An example is forecasting currency exchange rates. In \cite{rwMeese,rwCCE} the authors show that various econometric techniques and prediction models fail to consistently outperform Random Walk predictions in forecasting various exchange rates. This is the case because these signals are volatile, noisy and possess very little information structure but they don't \emph{on-average} vary a great deal. That is the day-to-day exchange rates \emph{on-average} don't vary very much so guessing the last known exchange rate is often a very good approximation to the new exchange rate. [[Not a huge fan of this wording.]] The Na\"ive forecasting scheme can be a great prediction strategy if the time series has very little forward information structure, adheres to a unimodal distribution and has very little variance. In this circumstance, forecasting the mean will on average get you in the ball park of the forecast without taking into account the lack of structure. The \naive forecasting scheme can be thought of as a filtering mechanism which compresses a signal onto a single point, effectively filtering out both complexity and structure. However, if very little predictive structure is present in a signal then this filtering mechanism effectively mitigates the complexity.

While the simplicity of these methods is advantageous on average when a time series exists on the high side of the complexity spectrum, their simplicity can also be a disadvantage and cause the methods to be inconsistent in their forecasting ability as a consequence of not being dynamic or taking into account the distribution or structure of the time series. Both of these methods fail spectacularly in two specific circumstances that lie on the fringes of their strengths. Fortunately, they somewhat parallel each other, i.e., the weakness of Random Walk and \naive are covered by the other primitive method. The situation in which Random Walk fails is when a time series changes drastically on a regular basis. The simplest example of this would be a period 2 time series which alternated between 0 and 1 at each time step. At every time step Random Walk would precisely predict out of phase, and as a result would accrue an error of 1 at every time step ($|p_i-X_{i,obs}|=1$. In comparison, the \naive method would forecast 0.5 at each time step and would thus accrue and error of 0.5 at every step. So on-average \naive would do twice as well as Random Walk, and as you could imagine a dynamic method would captured this periodicity  and exploited this predictive structure could do much much better, such a method is introduced in Section \ref{sec:lma}.

The \naive forecast fails when you have a signal which shifts between two or more regimes on a regular basis and the regimes are separated by a significant amount. In this case the \naive forecast will predict between the multiple regimes at every time step eventually accruing massive error. To see how Random Walk can cover this weakness imagine a time series which switches between more than one regime but stays in each regime for several time steps, this behaviour can be seen in some portions of the \svd trace seen in Figure \ref{fig:ts}. Random walk in this case could react to the regime shift within one time step, switch to the new Regime and begin forecasting nicely, where as \naive would just continue to predict between the modes of the multimodal distribution resulting in poor forecasting accuracy all of the time.

So clearly these primitive methods have their downfalls, but it is important to keep in mind that when a time series exists on the top of the complexity spectrum they have little predictive structure to exploit and often a primitive method like Random Walk or \naive will consistently out perform methods which rely on predictive structure. This is explored further in Section \ref{sec:accuracy} and \ref{sec:results}

\subsection{Linear Forecasting: Seasonal Autoregressive-integrated-moving average(ARIMA)}\label{sec:arima}
\begin{enumerate}
\item\cmark introduce method abstractly and practically, common method basically fitting a hyperplane to data removing seasonality and filtering for noise
\item\cmark define rigorously, all the backshift operator garbage

$$\Phi(B^m)\phi(B)(1-B^m)^D(1-B)^dX_i = c + \Theta(B^m)\theta(B)\epsilon_i$$

$$B^mX_i  = X_{i-m}$$
\item\cmark discuss that this method needs linear structure to work correctly
\item \cmark maybe talk about converging to zero and limits on prediction horizon
\end{enumerate}

@@

When there is evidence a generating process exhibits predictive structure an obvious strategy would be to fit a hyperplane to the dataset and then use this hyperplane to forecast new data points. This is a very old technique stemming back to Yule in 1927\cite{weigend93} with his invention of autoregressive forecasting, which forecasts the next time step through a weighted average of the past $p$ time-steps, i.e., $$p_i = \phi(B)X_{i,obs}.$$ Here $\phi(B)$ is a polynomial of degree $p$ and $B$ is the backshift operator, i.e., $$B^mX_i  = X_{i-m,obs}.$$
To improve this method further, especially in the presence of noisy data would be to allow for some wiggle room by adding a moving average of white noise to the model: $\theta(B)\epsilon_i$, where $\theta(B)$ is a polynomial of degree $q$ and $\{\epsilon_i\}\sim WN(0,\sigma^2)$. Finally if there is evidence of non-stationarity an initial differencing (integrated) is applied as an attempt to remove that non-stationarity: $(1-B^d)X_i$. Combining these three give us a nonseasonal  ARIMA$(p,d,q)$ model:  $$\phi(B)(1-B)^dX_i = \theta(B)\epsilon_i.$$ Where $p$, $d$ and $q$ correspond to the orders of the autoregressive, integrated and moving average models respectively. If periodic structure is present a seasonal ARIMA$(p,d,q)(P,D,Q)_m$ is often used and is defined by:$$\Phi(B^m)\phi(B)(1-B^m)^D(1-B)^dX_i = \Theta(B^m)\theta(B)\epsilon_i.$$ Here $\Phi$ and $\Theta$ are polynomials of degree P and Q respectively, and $m$ is the seasonal frequency. Forecasting is then done in two steps, first the integrated step is performed to remove non-stationarity. $$\hat{X_i} = (1-B^m)^D(1-B)^d X_{i,obs}$$. Then forecasts are made by evaluating $$\Phi(B^m)\phi(B)\hat{X_i} = \Theta(B^m)\theta(B)\epsilon_i.$$

@@

ARIMA forecasting is a time tested modelling process which uses a global linear fit and adjustments for seasonality, non-stationarity and noise. If a generating process fits into one of these categories then this model is a great short term predictor. If information is instead being transmitted in a nonlinear way this global linear fit falls apart. One other weakness of this method is that if long prediction horizons are used the forecast is guaranteed to converge to the mean after some number of predictions. For this reason, we limit all methods to doing multiple 1-step predictions as to not handicap this method.

\subsection{Nonlinear Forecasting: Lorenz Method of Analogues (LMA)}\label{sec:lma}

@@

 \subsubsection{Reconstructing hidden dynamics}

@@

%% can cut for space if need be:
The coordinates of each point on this plot are differently delayed
elements of the \col instructions per cycle time series
$X_{i,obs}$: that is, $X_{i,obs}$ on the first axis, $X_{i+\tau,obs}$ on the second,
$X_{i+2\tau,obs}$ on the third, and so on.
Structure in these kinds of plots---clearly visible in
Figure~\ref{fig:embedding}---is an indication of
determinism%\footnote{A deeper analysis of
  %Figure~\ref{fig:embedding}---as alluded to on the previous
  %page---supports that diagnosis, confirming the presence of a chaotic
  %attractor in these cache-miss dynamics, with largest Lyapunov
  %exponent $\lambda_1 = 8000 \pm 200$ instructions, embedded in a
  %12-dimensional reconstruction space \cite{mytkowicz09}.}
  .  That structure can be used to build a forecast model.
 \subsubsection{LMA: Using dynamics in forecasting}

@@

\subsection{Prediction Accuracy: Mean Absolute Scaled Error (MASE)}
\label{sec:accuracy}

>>>>

\section{Modeling {\color{blue} EDITABLE}}\label{sec:model}

@@

\subsection{Primitive Strategies to combat complexity: Random Walk and \naive {\color{blue} EDITABLE}}\label{sec:simple}
In this section we cover two very simple forecast models: Random Walk and \naive. With a Random Walk predictor to forecast at time $i$ we simply use the last observed measurement: $$p_i = X_{i-1,obs}.$$ The \naive  prediction strategy is to forecast the rolling average up to that point. i.e., the prediction $p_i$ is simply the average of all observed points up to that point: $$p_i = \sum_{j=1}^{i-1}\frac{X_{j,obs}}{i-1}.$$

While both of these methods are simplistic they are not without merit. When time series possess very little predictive structure,i.e., they exist on the very top of the complexity spectrum these two methods can sometimes be \emph{on-average} very good choices for a forecasting method, as we will see in our results. An example is forecasting currency exchange rates. In \cite{rwMeese,rwCCE} the authors show that various econometric techniques and prediction models fail to consistently outperform Random Walk predictions in forecasting various exchange rates. This is the case because these signals are volatile, noisy and possess very little information structure but they don't \emph{on-average} vary a great deal. That is the day-to-day exchange rates \emph{on-average} don't vary very much so guessing the last known exchange rate is often a very good approximation to the new exchange rate. [[Not a huge fan of this wording.]] The Na\"ive forecasting scheme can be a great prediction strategy if the time series has very little forward information structure, adheres to a unimodal distribution and has very little variance. In this circumstance, forecasting the mean will on average get you in the ball park of the forecast without taking into account the lack of structure. The \naive~ forecasting scheme can be thought of as a filtering mechanism which compresses a signal onto a single point, effectively filtering out both complexity and structure. However, if very little predictive structure is present in a signal then this filtering mechanism effectively mitigates the complexity.

While the simplicity of these methods is advantageous on average when a time series exists on the high side of the complexity spectrum, their simplicity can also be a disadvantage and cause the methods to be inconsistent in their forecasting ability as a consequence of not being dynamic or taking into account the distribution or structure of the time series. Both of these methods fail spectacularly in two specific circumstances that lie on the fringes of their strengths. Fortunately, they somewhat compliment each other, i.e., the weakness of Random Walk and \naive~ are covered by the other primitive method.

The situation in which Random Walk fails is when a time series changes drastically on a regular basis. The simplest example of this would be a period 2 time series which alternated between 0 and 1 at each time step. At every time step Random Walk would precisely predict out of phase, and as a result would accrue an error of 1 at every time step ($|p_i-X_{i,obs}|=1$. In comparison, the \naive method would forecast 0.5 at each time step and would thus accrue an error of 0.5 at every step. So on-average \naive would do twice as well as Random Walk, and as you could imagine a dynamic method would captured this periodicity  and exploited this predictive structure could do much much better, such a method is introduced in Section \ref{sec:lma}.

The \naive forecast fails when you have a signal which shifts between two or more regimes on a regular basis and the regimes are separated by a significant amount. In this case the \naive forecast will predict between the regimes at every time step eventually accruing massive error. To see how Random Walk can cover this weakness imagine a time series which switches between more than one regime but stays in each regime for several time steps, e.g., a regularly sampled noisy square wave, this behavior can be seen in some portions of the \svd trace seen in Figure \ref{fig:svd}. Random walk in this case could react to the regime shift within one time step, switch to the new Regime and begin forecasting nicely, where as \naive would just continue to predict between the modes of the multimodal distribution resulting in poor forecasting accuracy all of the time.

So clearly these primitive methods have their downfalls, but it is important to keep in mind that when a time series exists on the top of the complexity spectrum they have little predictive structure to exploit and often a primitive method like Random Walk or \naive~ will consistently out perform methods which rely on predictive structure. This is explored further in Section \ref{sec:accuracy} and \ref{sec:results}

\subsection{Linear Forecasting: Seasonal Autoregressive-integrated-moving average(ARIMA){\color{blue} EDITABLE}}\label{sec:arima}
%\begin{enumerate}
%\item\cmark introduce method abstractly and practically, common method basically fitting a hyperplane to data removing seasonality and filtering for noise
%\item\cmark define rigorously, all the backshift operator garbage

%$$\Phi(B^m)\phi(B)(1-B^m)^D(1-B)^dX_i = c + \Theta(B^m)\theta(B)\epsilon_i$$

%$$B^mX_i  = X_{i-m}$$
%\item\cmark discuss that this method needs linear structure to work correctly
%\item \cmark maybe talk about converging to zero and limits on prediction horizon
%\end{enumerate}

@@

When there is evidence a generating process exhibits predictive structure an obvious strategy would be to fit a hyperplane to the dataset and then use this hyperplane to forecast new data points. This is a very old technique stemming back to Yule in 1927\cite{weigend93} with his invention of autoregressive forecasting, which forecasts the next time step through a weighted average of the past $p$ time-steps, i.e., $$p_i = \phi(B)X_{i,obs}.$$ Here $\phi(z)$ is a polynomial of degree $p$ and $B$ is the backshift operator, i.e., $$B^mX_i  = X_{i-m,obs}.$$
To improve this method further--especially in the presence of noisy data---one could allow for some wiggle room by adding a moving average of white noise to the model: $\theta(B)\epsilon_i$, where $\theta(z)$ is a polynomial of degree $q$ and $\{\epsilon_i\}\sim WN(0,\sigma^2)$. Finally if there is evidence of non-stationarity an initial differencing (integrated) is applied as an attempt to remove that non-stationarity: $(1-B^d)X_i$. Combining these three give us a nonseasonal  ARIMA$(p,d,q)$ model:  $$\phi(B)(1-B)^dX_i = \theta(B)\epsilon_i.$$ Where $p$, $d$ and $q$ correspond to the orders of the autoregressive, integrated and moving average models respectively. If periodic structure is present a seasonal ARIMA$(p,d,q)(P,D,Q)_m$ is often used and is defined by:$$\Phi(B^m)\phi(B)(1-B^m)^D(1-B)^dX_i = \Theta(B^m)\theta(B)\epsilon_i.$$ Here $\Phi(z)$ and $\Theta(z)$ are polynomials of degree P and Q respectively, and $m$ is the seasonal frequency. Forecasting is then done in two steps, first the integrated step is performed to remove non-stationarity. $$\hat{X_i} = (1-B^m)^D(1-B)^d X_{i,obs}.$$ Then forecasts are made by evaluating $$\Phi(B^m)\phi(B)\hat{X_i} = \Theta(B^m)\theta(B)\epsilon_i.$$

@@

ARIMA forecasting is a time tested modelling process which uses a global linear fit and adjustments for seasonality, non-stationarity and noise. If a generating process fits into this category then this model is a great short term predictor. If information is instead being transmitted in a nonlinear way this global linear fit falls apart. One other weakness of this method is that if long prediction horizons are needed the forecast is guaranteed to converge to the mean after some number of predictions (dependent on model order). For this reason, we limit all methods to doing multiple 1-step predictions as to not handicap this method.

\subsection{Nonlinear Forecasting: Lorenz Method of Analogues (LMA){\color{red} NOT EDITABLE}}\label{sec:lma}

@@

 \subsubsection{Reconstructing hidden dynamics {\color{red} NOT EDITABLE}}

@@

%% can cut for space if need be:
The coordinates of each point on this plot are differently delayed
elements of the \col instructions per cycle time series
$X_{i,obs}$: that is, $X_{i,obs}$ on the first axis, $X_{i+\tau,obs}$ on the second,
$X_{i+2\tau,obs}$ on the third, and so on.
Structure in these kinds of plots---clearly visible in
Figure~\ref{fig:embedding}---is an indication of
determinism%\footnote{A deeper analysis of
  %Figure~\ref{fig:embedding}---as alluded to on the previous
  %page---supports that diagnosis, confirming the presence of a chaotic
  %attractor in these cache-miss dynamics, with largest Lyapunov
  %exponent $\lambda_1 = 8000 \pm 200$ instructions, embedded in a
  %12-dimensional reconstruction space \cite{mytkowicz09}.}
  .  That structure can be used to build a forecast model.
 \subsubsection{LMA: Using dynamics in forecasting {\color{red} NOT EDITABLE}}

@@

\subsection{Prediction Accuracy: Mean Absolute Scaled Error (MASE) {\color{red} NOT EDITABLE}}
\label{sec:accuracy}

### 2014-03-04 results.tex
editing

 \section{Predictability, Complexity, and Permutation Entropy} % (fold)
 \label{sec:results}
% I changed the title of this section because there are lots of
% results in the modeling section too.  The results HERE are about the
% relationship between predictability and WPE

@@

The WPE of \svdfive ($0.677 \pm 0.006$) is higher than that of {\tt
  col\_major}.  This indicates that the rate of forward information
transfer of the underlying process is lower, but that time-series data
from this system still contain a significant amount of structure that
can, in theory, be leveraged to predict the future course of the time
series.  As mentioned above, though, in the discussion of {\tt
  col\_major}, the effectiveness of any prediction strategy will
depend on how well it leverages the available information; methods
that use linear models, for instance, may fail when applied to
nonlinear processes.
% 
% (This is the basis for the second conjecture above.)  
% 
The match between model and task is an issue in the \svdfive
experiments as well.  The MASE scores of the \naive and ARIMA
predictions for this system were $20.870 \pm 0.192$ and $2.370 \pm
0.051$, respectively: that is, 20.87 and 2.37 times worse than a
simple random walk forecast of the same signals\footnote{The \naive
  MASE score is large because the amount of variance in this signal is
  high, which makes guessing the mean a particularly bad strategy.
  The same thing is true of the \svdthree signal.}.  Again, the
positions of these points on a WPE vs. MASE plot---significantly below
and to the right of the dashed line---should suggest to a practitioner
that there is more structure in this signal than the corresponding
method is able to leverage, and that a different method might do
better.  Indeed, for \svdfive, the LMA method produces a MASE score of
$ 0.718\pm 0.048 $ and a cluster of results that is much closer to the
dashed line on the WPE-MASE plot.  Again, this validates our second
conjecture: the LMA method can capture and reproduce the way in which
the \svdfive system processes information, but the \naive and ARIMA
prediction methods cannot.

>>>>

 \section{Predictability, Complexity, and Permutation Entropy {\color{blue}  EDITABLE}} % (fold)
 \label{sec:results}
% I changed the title of this section because there are lots of
% results in the modeling section too.  The results HERE are about the
% relationship between predictability and WPE

@@

The WPE of \svdfive ($0.677 \pm 0.006$) is higher than that of {\tt
  col\_major}.  This indicates that the rate of forward information
transfer of the underlying process is lower, but that time-series data
from this system still contain a significant amount of structure that
can, in theory, be leveraged to predict the future course of the time
series.  As mentioned above, though, in the discussion of {\tt
  col\_major}, the effectiveness of any prediction strategy will
depend on how well it leverages the available information; methods
that use linear models, for instance, may fail when applied to
nonlinear processes.
% 
% (This is the basis for the second conjecture above.)  
% 
The match between model and task is an issue in the \svdfive
experiments as well.  The MASE scores of the \naive and ARIMA
predictions for this system were $20.870 \pm 0.192$ and $2.370 \pm
0.051$, respectively: that is, 20.87 and 2.37 times worse than a
simple random walk forecast of the same signals\footnote{The \naive
  MASE score is large because the amount of variance in this signal is
  high, which makes guessing the mean a particularly bad strategy.
  The same thing is true of the \svdthree signal.[[Joshua, it isn't the variance it is the strength in the bimodal distribution and only predicting in one mode of the distribution...]]}.  Again, the
positions of these points on a WPE vs. MASE plot---significantly below
and to the right of the dashed line---should suggest to a practitioner
that there is more structure in this signal than the corresponding
method is able to leverage, and that a different method might do
better.  Indeed, for \svdfive, the LMA method produces a MASE score of
$ 0.718\pm 0.048 $ and a cluster of results that is much closer to the
dashed line on the WPE-MASE plot.  Again, this validates our second
conjecture: the LMA method can capture and reproduce the way in which
the \svdfive system processes information, but the \naive and ARIMA
prediction methods cannot.

### 2014-03-04 wpe.tex
adding experimental methods/collection section and updating
editing

\section{Measuring Structural Complexity}\label{sec:meaComplex}

@@

\item make sure it is clear and we justify that detecting and characterizing structural complexity is hard for real-valued noisy time series.
\item Quantifying structured and unstructured complexity is nontrivial in the case of real-valued noisy time series but WPE does this. [[talk about again here but justify in information theory section]]
\item justify and explain the regime split in \svd 
\end{enumerate}

>>>>

\section{Measuring Structural Complexity {\color{red} NOT EDITABLE}}\label{sec:meaComplex}

@@

\item make sure it is clear and we justify that detecting and characterizing structural complexity is hard for real-valued noisy time series.
\item Quantifying structured and unstructured complexity is nontrivial in the case of real-valued noisy time series but WPE does this. [[talk about again here but justify in information theory section]]
\item justify and explain the regime split in \svd {\color{red} I(joshua) moved this justification in the experimental methods section, it just needs to be explored here by applying windowed wpe and explaining that method}
\end{enumerate}
